---
title: "Practical Machine Learning Course Project"
subtitle: "Practical Machine Learning project"
author: "Elmer Ore"
output: html_document
---

# Executive Summary

In this report I aim to replicate the findings of the Human Activity Recognition experiment and be able to identify how well the 6 partifipants did the dumbbell bicepts curl vs. common mistakes being recorded.  By analyzing the data provided I exclude some predictors that did not have a relationship wih the outcome or predictors with high number of null or empty values.  For this classification project I created a recipe to prepare the data and choosed to use random forest as the model to solve the problem.  From this model **I found that 26 predictors shows the highest accuracy of 99.57% for Out-of-bag error.**  Finally, I included the list of the top 20 variables that most affects the final model.

## Exploratory Analysis

- Identified mostly null or empty variables, I decided 97% as a cutoff value.
- Also identified fields not useful in predicting how the exercise was done.  Decided to remove the variables: unique sequential number, user name, timestamps, and window information.

## Reproducible Research

- Decided to use recipes, to make is clearer the steps to reproduce the analysis.
- Removed the fields identified in the exploratory analysis.
- Exclude near zero variace and variable correlated.
- Converted all factor variables into dummy variables, except the outcome.
- Center and scale the rest of variables.
\newline


```{r setup, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}


```

## Model Selection

I decided to use 10-fold cross validation due to the size of the training set and number of variables available. 
In terms of model, I decided to use the random forests model. 

```{r pressure, echo=FALSE}
plot(pressure)
```

### Model Validation

To validate if the 3 variables identified in the AIC methos are really necessary I did an "Analysis of Variance" (ANOVA) test using nested models. The results show that the 3rd model has a significant reduction of RSS and better model fits. See Appendix #3 for details.
\newline

## Conclusions

The model (see summary on Appendix #2) shows:

- A manual transmission car produces in average 2.9 MPG higher performance than an automatic car with all other variables being constant with 95% accuracy.
- Not statisticall signficant effect of autmatic cars on MPG performance (p-value = 17.79%).
- An extra 1000 lbs in weight results in -3.9 MPG decrease in performance with all other variables being constant with 95% accuracy. 
- An extra second on going through 1/4 mile results in an additional 1.2 MPG with all other variables being constant with 95% accuracy.
- The model explains 84.97% variability of MPG.
\newpage

## References

- The data used for this project comes from <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har>.
